# Lesson 5a: Introduction to ggplot2

```{r setup-5a, include=FALSE}
# clean up saved items that may carry over from previous lessons
rm(list = ls())

knitr::opts_chunk$set(
  fig.align = "center",
  fig.height = 3.5,
  fig.width = 6,
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

Being able to create visualizations (graphical representations) of data is a key step in being able to communicate information and findings to others. In this lesson, you will learn to use the **ggplot2** library to make beautiful plots or charts of your data. Although R does provide built-in plotting functions, the **ggplot2** library implements the *Grammar of Graphics*. This makes it particularly effective for describing how visualizations should represent data, and has turned it into the preeminent plotting library in R. Learning this library will allow you to make nearly any kind of (static) data visualization, customized to your exact specifications.

## Learning objectives

By the end of this lesson you will be able to:

* Plot data using various geometric shapes and statistical transformations.
* Manipulate and control the aesthetics of your plots (i.e. colors, scales).
* Put together various ggplot parameters to produce a refined plot with a title, subtitle, proper axis titles, etc.

## Prerequisites

To reproduce the code throughout this lesson you will need to load the **ggplot2** package along with a couple others. 

```{r}
library(ggplot2)          # for plotting capabilities
library(completejourney)  # for data
library(dplyr)            # for additional data wrangling
```

This lesson will also use the Complete Journey data; but before we start we'll join the transactions data with the products and household demographics data:

```{r}
df <- transactions_sample %>%
   inner_join(products) %>%
   inner_join(demographics)

glimpse(df)
```


## Grammar of Graphics

Just as the grammar of language helps us construct meaningful sentences out of words, the *Grammar of Graphics* helps us to construct graphical figures out of different visual elements. This grammar gives us a way to talk about parts of a plot: all the circles, lines, arrows, and words that are combined into a diagram for visualizing data. [Originally developed by Leland Wilkinson](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448), the Grammar of Graphics was [adapted by Hadley Wickham](https://ggplot2-book.org/) to describe the components of a plot, including

* the **data** being plotted
* the **geometric objects** (circles, lines, etc.) that appear on the plot
* a set of mappings from variables in the data to the **aesthetics** (appearance) of the geometric objects
* a **statistical transformation** used to calculate the data values used in the plot
* a **position adjustment** for locating each geometric object on the plot
* a **scale** (e.g., range of values) for each aesthetic mapping used
* a **coordinate system** used to organize the geometric objects
* the **facets** or groups of data shown in different plots

Wickham further organizes these components into layers, where each layer has a single *geometric object*, *statistical transformation*, and *position adjustment*. Following this grammar, you can think of each plot as a set of layers of images, where each image’s appearance is based on some aspect of the data set.

All together, this grammar enables us to discuss what plots look like using a standard set of vocabulary. And similar to how tidyr and dplyr provide efficient data transformation and manipulation, **ggplot2** provides more efficient ways to create specific visual images.

## The Basics

In order to create a plot, you:

1. Call the `ggplot()` function which creates a blank canvas
2. Specify **aesthetic mappings**, which specifies how you want to map variables to visual aspects. In this case we are simply mapping the `quantity` and `sales_value` variables to the x- and y-axes.
3. You then add new layers that are geometric objects which will show up on the plot. In this case we add `geom_point` to add a layer with points (dots) elements as the geometric shapes to represent the data.

```{r first-plot-scatterplot, eval=FALSE}
# create canvas (left)
ggplot(df)

# variables of interest mapped (middle)
ggplot(df, aes(x = quantity, y = sales_value))

# data plotted (right)
ggplot(df, aes(x = quantity, y = sales_value)) +
   geom_point()
```

```{r first-plot-scatterplot-rendered, echo=FALSE, fig.width=11}
library(gridExtra)

# create canvas
p1 <- ggplot(df) + ggtitle("Canvas")

# variables of interest mapped
p2 <- ggplot(df, aes(x = quantity, y = sales_value)) +
  ggtitle("Canvas + variables mapped to axes")

# data plotted
p3 <- ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point() +
  ggtitle("Data plotted")

grid.arrange(p1, p2, p3, nrow=1)
```

```{block, type="note"}
When you add the geom layer you use the addition (+) operator. As you add new layers you will always use + to add onto your visualization.
```

Just as we've seen in the previous lessons, we can continue to use the pipe operator with ggplot. For example, say we want to only plot transactions where the quantity of items purchased was over 100. We can simply apply a `filter()` statement and then pipe the results directly into ggplot.

```{block, type="warning"}
Just remember to use the + operator to add new layers to ggplot. It is very common for people to pipe into ggplot and then continue to use the %>% operator to add new ggplot layers. This will result in an error!
```

```{r}
df %>%
   filter(quantity > 100) %>%
   ggplot(aes(x = quantity, y = sales_value)) +
   geom_point()
```

### Knowledge check

```{block, type='todo'}
1. Filter for all transactions in the produce `department` and plot the `quantity` on the x-axis and `sales_value` on the y-axis.
2. Filter for all transactions in the meat `department`s and plot the `quantity` on the x-axis and `sales_value` on the y-axis.
3. Filter for all transactions in the fuel `department`s and plot the `quantity` on the x-axis and `sales_value` on the y-axis.
4. What trends do you notice between the above departments?
```

## Aesthetic Mappings

The ___aesthetic mappings___ take properties of the data and use them to influence visual characteristics, such as *position*, *color*, *size*, *shape*, or *transparency*. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics for a plot are specified in the `aes()` function call (later in this lesson you will see that each `geom` layer can have its own `aes` specification). For example, we can add a mapping from the variable `household_size` to a color characteristic:

```{r scatterplot-by-store, echo=TRUE}
df %>%
   ggplot(aes(x = quantity, y = sales_value, color = household_size)) +
   geom_point()
```


Note that using the `aes()` function will cause the visual channel to be based on the data specified in the argument. For example, using `aes(color = "blue")` won’t cause the geometry’s color to be “blue”, but will instead cause the visual channel to be mapped from the vector `c("blue")` — as if we only had a single type of engine that happened to be called “blue”. If you wish to apply an aesthetic property to an entire geometry, you can set that property as an argument within the `geom_xxx_()` method, outside of the `aes()` call:

```{r scatterplot-specifying-all-points-as-blue, fig.width=5}
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point(color = "blue")
```

### Knowledge check

```{block, type='todo'}
1. Filter for all transactions in the grocery `department` (`department == "GROCERY"`) and plot the `quantity` on the x-axis and `sales_value` on the y-axis.
2. Color all the points in the above plot red. Now color the points in #1 based on the `household_size` variable.
3. Use `size` to change the size of all points to 0.5. Now map the `household_size` variable to size within `aes()`.
```


## Specifying Geometric Shapes

Building on these basics, **ggplot2** can be used to build almost any kind of plot you may want. These plots are declared using functions that follow from the *Grammar of Graphics*.

The most obvious distinction between plots is what **geometric objects** (`geoms`) they include. **ggplot2** supports a number of different types of `geoms`, including:

* `geom_point` for drawing individual points (e.g., a scatter plot)
* `geom_line` for drawing lines (e.g., for a line charts)
* `geom_smooth` for drawing smoothed lines (e.g., for simple trends or approximations)
* `geom_bar` for drawing bars (e.g., for bar charts)
* `geom_histogram` for drawing binned values (e.g. a histogram)
* `geom_polygon` for drawing arbitrary shapes
* `geom_map` for drawing polygons in the shape of a map! (You can access the data to use for these maps by using the `map_data()` function).

Each of these geometries will leverage the aesthetic mappings supplied although the specific visual properties that the data will map to will vary. For example, you can map data to the `shape` of a `geom_point` (e.g., if they should be circles or squares), or you can map data to the `linetype` of a `geom_line` (e.g., if it is solid or dotted), but not vice versa.

Almost all `geoms` require an `x` and `y` mapping at the bare minimum.

```{r, eval=FALSE}
# Left column: x and y mapping needed!
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point()

ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_smooth()

# Right column: no y mapping needed!
ggplot(df, aes(x = household_size)) +
  geom_bar() 

ggplot(df, aes(x = sales_value)) +
  geom_histogram()
```


```{r x-and-y-mapping-scatterplot, fig.width=10, fig.height=8, echo=FALSE}

# Left column: x and y mapping needed!
p1 <- ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point() +
  ggtitle("x and y mapping needed!")

p2 <- ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_smooth()

# Right column: no y mapping needed!
p3 <- ggplot(df, aes(x = household_size)) +
  geom_bar() +
  ggtitle("no y mapping needed!")

p4 <- ggplot(df, aes(x = sales_value)) +
  geom_histogram()


graphs<-grid.arrange(p1, p3, p2, p4, ncol = 2)
```

What makes this really powerful is that you can add ___multiple___ geometries to a plot, thus allowing you to create complex graphics showing multiple aspects of your data.

```{r scatterplot-with-smoothed-line, fig.width=5}
# plot with both points and smoothed line
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point() +
  geom_smooth()
```


Of course the aesthetics for each `geom` can be different, so you could show multiple lines on the same plot (or with different colors, styles, etc). It’s also possible to give each `geom` a different data argument, so that you can show multiple data sets in the same plot.

For example, we can plot both points and a smoothed line for the same `x` and `y` variable but specify unique colors within each `geom`:

```{r scatterplot-with-smoothed-line-and-specified-color, fig.width=5}
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point(color = "blue") +
  geom_smooth(color = "red")
```


So as you can see if we specify an aesthetic within `ggplot` it will be passed on to each `geom` that follows. Or we can specify certain aesthetics within each `geom`, which allows us to only show certain characteristics for that specific layer (i.e. `geom_point`).

```{r scatter-plot-with-colored-lines-by-division, eval=FALSE}
# color aesthetic passed to each geom layer (left)
ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
  geom_point() +
  geom_smooth(se = FALSE)

# color aesthetic specified for only the geom_point layer (right)
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point(aes(color = household_size)) +
  geom_smooth(se = FALSE)
```

```{r scatter-plot-with-colored-lines-by-division2, echo=FALSE, fig.width=12}
# color aesthetic passed to each geom layer
p1 <- ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
  geom_point() +
  geom_smooth(se = FALSE)

# color aesthetic specified for only the geom_point layer
p2 <- ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point(aes(color = household_size)) +
  geom_smooth(se = FALSE)

grid.arrange(p1, p2, nrow = 1)
```

### Knowledge check

```{block, type='todo'}
1. Filter all transactions where `department == "PRODUCE"`. Use a boxplot to plot the `sales_value` across each household composition (`household_comp`). Does it look like the household composition impacts the amount spent on produce?
2. Using the same filtered data as in #1, create a scatter plot to assess `quantity` versus `sales_value` and color the points based on household composition.
3. Plot a histogram of the `retail_disc`ount values based on all transactions.
```

## Statistical Transformations

If you look at the bar chart below displaying the frequency of store IDs, you’ll notice that the the y axis was defined for us as the _count_ of observations for each **household_size**. This count isn’t part of the data set (it’s not a column in the df data set), but is instead a ___statistical transformation___ that the `geom_bar` automatically applies to the data. In particular, it applies the `stat_count` transformation.


```{r statistical-transformation}
ggplot(df, aes(x = household_size)) +
  geom_bar() 
```


**ggplot2** supports many different statistical transformations. For example, the “identity” transformation will leave the data “as is”. You can specify which statistical transformation a `geom` uses by passing it as the `stat` argument. For example, consider our data already had the count as a variable:

```{r show-dplyr-count-function}
household_size_count <- count(df, household_size)
household_size_count
```


We can use `stat = "identity"` within `geom_bar` to plot our bar height values to this variable. Also, note that we now include n for our y variable:

```{r counts-bar-plot}
ggplot(household_size_count, aes(x = household_size, y = n)) +
  geom_bar(stat = "identity")
```


We can also call `stat_` functions directly to add additional layers. For example, below is a graph that visualizes the relationship between the total sales value for each basket (aka checkout) across the different income levels. We create a scatterplot of total sales value for each for each basket and then use `stat_summary` to plot the median total sales value. It appears that there is a slight increase in median checkout sales value as income increases. 

```{r stat-transformation-mean, fig.width=8}
df %>%
   group_by(income, basket_id) %>%
   summarize(total_sale = sum(sales_value)) %>%
   ggplot(aes(income, total_sale)) + 
   geom_point(color = "blue") + 
   stat_summary(fun = "median", geom = "point", color = "red", size = 2)  
```

### Knowledge check

```{block, type='todo'}
1. Filter our transactions for only those items in the meat `department` and use `geom_bar()` to plot the number of transactions for each `product_category`.
2. Filter our transactions for only those items in the meat `department` and you did above but now compute the number of transactions for each `product_category` using `filter::count()`. Use `geom_count()` to create the same plot as you got in #1.
3. Create a scatter plot comparing `retail_disc` to `sales_value`. Now, instead of using `geom_point()` use `stat_summary_bin(fun = "mean", geom = "bar", orientation = 'y')`. What does this tell you about the relationship between `retail_disc` and `sales_value`?
```

## Position Adjustments

In addition to a default statistical transformation, each `geom` also has a default ___position adjustment___ which specifies a set of “rules” as to how different components should be positioned relative to each other. This position is noticeable in a `geom_bar` if you map a different variable to the color visual characteristic. Below is a bar chart that uses color to visualize the proportion of brand purchases (Private represents the stores special private brand and National represents commonly available brands).

```{r position-adjustments}
ggplot(df, aes(x = household_size, fill = brand)) +
   geom_bar()
```

The `geom_bar` by default uses a position adjustment of `"stack"`, which makes each rectangle’s height proprotional to its value and stacks them on top of each other. We can use the `position` argument to specify what position adjustment rules to follow:


```{r bar-chart-positions, eval=FALSE}
# position = "dodge": values next to each other
ggplot(df, aes(x = household_size, fill = brand)) + 
  geom_bar(position = "dodge")

# position = "fill": percentage chart
ggplot(df, aes(x = household_size, fill = brand)) + 
  geom_bar(position = "fill")
```

```{r bar-chart-positions2, echo=FALSE, fig.width=12, fig.height=4}
# position = "dodge": values next to each other
p1 <- ggplot(df, aes(x = household_size, fill = brand)) + 
  geom_bar(position = "dodge") +
  ggtitle("position = 'dodge'")

# position = "fill": percentage chart
p2 <- ggplot(df, aes(x = household_size, fill = brand)) + 
  geom_bar(position = "fill") +
  ggtitle("position = 'fill'")

grid.arrange(p1, p2, nrow = 1)
```

```{block, type="tip"}
Check the documentation for each particular `geom` to learn more about its positioning adjustments.
```

### Knowledge check

```{block, type='todo'}
1. Create a bar plot that visualizes the number of transactions across all `weeks`.
2. Now add a `fill` parameter to assess if the ratio of `brand` was consistent across these weeks.
```

## Managing Scales

Whenever you specify an aesthetic mapping, `ggplot` uses a particular ___scale___ to determine the range of values that the data should map to. Thus when you specify

```r
# color the data by household_size
ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
   geom_point()
```

`ggplot` automatically adds a scale for each mapping to the plot:


```r
# same as above, with explicit scales
ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
   geom_point()
   scale_x_continuous() +
   scale_y_continuous() +
   scale_colour_discrete()
```
  
Each scale can be represented by a function with the following name: `scale_`, followed by the name of the aesthetic property, followed by an `_` and the name of the scale. A `continuous` scale will handle things like numeric data (where there is a _continuous_ set of numbers), whereas a `discrete` scale will handle things like colors (since there is a small list of _distinct_ colors). If you are trying to use a `continuous` scale, you must make sure the data is in numeric format and not as a factor or character.

While the default scales will work fine, it is possible to explicitly add different scales to replace the defaults. For example, you can use a scale to change the direction of an axis:

```{r flip-axes}
# total area size and spend relationship, ordered in reverse
ggplot(df, aes(x = quantity, y = sales_value)) +
   geom_point() +
   scale_x_reverse() +
   scale_y_reverse()
```
  
  
Similarly, you can use `scale_x_log10()` and `scale_x_sqrt()` to transform your scale. You can also use `scales` to format your axes. For example, you can make the y-axis have a percent scale.

```{r bar-y-axis-percentage-format}
ggplot(df, aes(x = household_size, fill = brand)) + 
  geom_bar(position = "fill") +
  scale_y_continuous(breaks = seq(0, 1, by = .2), labels = scales::percent)
```  
  

  A common parameter to change is which set of colors to use in a plot. While you can use the default coloring, a more common option is to leverage the pre-defined palettes from [colorbrewer.org](colorbrewer.org). These color sets have been carefully designed to look good and to be viewable to people with certain forms of color blindness. We can leverage color brewer palletes by specifying the `scale_color_brewer()` function, passing the pallete as an argument. 

```{r color-brewer-palette-example, eval=FALSE}
# left: default color brewer
ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
  geom_point() +
  scale_color_brewer()

# right: specifying color palette
ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
  geom_point() +
  scale_color_brewer(palette = "Set3")
```
  
```{r color-brewer-palette-example2, echo=FALSE, fig.width=12}
# default color brewer
p1 <- ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
  geom_point() +
  scale_color_brewer()

# specifying color palette
p2 <- ggplot(df, aes(x = quantity, y = sales_value, color = household_size)) +
  geom_point() +
  scale_color_brewer(palette = "Set3")

grid.arrange(p1, p2, nrow=1)
```  
  
Note that you can get the palette name from the [colorbrewer website](http://colorbrewer2.org/) by looking at the scheme query parameter in the URL. Or see the diagram [here](https://bl.ocks.org/mbostock/5577023) and hover the mouse over each palette for the name.

You can also specify _continuous_ color values by using a [gradient scale](https://ggplot2.tidyverse.org/reference/scale_gradient.html), or [manually](https://ggplot2.tidyverse.org/reference/scale_manual.html) specify the colors you want to use as a named vector.

### Knowledge check

```{block, type='todo'}
1. Create a scatter plot that plots `quantity` vs `sales_value` across all transactions. View this plot. Now transform the x-axis to a log transformation using `scale_x_log10()`. What new insights does this provide? Now add `labels = scales::comma` inside `scale_x_log10()`. How does this change the formatting?
2. Create a scatter plot that plots `sales_value` vs `retail_disc` across all transactions. View this plot. Now transform the x-axis ***and*** y-axis to a log transformation using `scale_x_log10()` and `scale_y_log10()`. What new insights does this provide? Can you figure out how to change the x and y-axes to be formatted as dollar units?
```

## Coordinate Systems

The next term from the *Grammar of Graphics* that can be specified is the ___coordinate system___. As with scales, coordinate systems are specified with functions that all start with `coord_` and are added as a layer. There are a number of different possible coordinate systems to use, including:

* `coord_cartesian` the default [cartesian coordinate system](https://en.wikipedia.org/wiki/Cartesian_coordinate_system), where you specify x and y values (e.g. allows you to zoom in or out).
* `coord_flip` a cartesian system with the x and y flipped
* `coord_fixed` a cartesian system with a “fixed” aspect ratio (e.g., 1.78 for a “widescreen” plot)
* `coord_polar` a plot using [polar coordinates](https://en.wikipedia.org/wiki/Polar_coordinate_system)
* `coord_quickmap` a coordinate system that approximates a good aspect ratio for maps. See documentation for more details.

```{r coordinate, eval=FALSE}
# zoom in with coord_cartesian (left)
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0, 50))

# flip x and y axis with coord_flip (right)
ggplot(df, aes(x = household_size)) +
  geom_bar() +
  coord_flip()
```
  
```{r coordinate2, echo=FALSE, fig.width=10}
# zoom in with coord_cartesian
p1 <- ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0, 50))

# flip x and y axis with coord_flip
p2 <- ggplot(df, aes(x = household_size)) +
  geom_bar() +
  coord_flip()

grid.arrange(p1, p2, nrow=1)
```  

### Knowledge check

```{block, type='todo'}
Produce a regular bar plot based on the `department` variable. Now change this plot from a vertical bar plot to a horizontal bar plot using `coord_flip()`.
```

## Facets

___Facets___ are ways of grouping a data plot into multiple different pieces (subplots). This allows you to view a separate plot for each value in a categorical variable. You can construct a plot with multiple facets by using the `facet_wrap()` function. This will produce a “row” of subplots, one for each categorical variable (the number of rows can be specified with an additional argument):

```{r facets1, fig.width=12, fig.height=4}
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point() +
  facet_grid(~ marital_status)
```  
  
You can also use `facet_grid` to facet your data by more than one categorical variable. Note that we use a tilde (`~`) in our facet functions. With `facet_grid` the variable to the left of the tilde will be represented in the rows and the variable to the right will be represented across the columns.

```{r facet-grid-plot, fig.width=12, fig.height=8}
ggplot(df, aes(x = quantity, y = sales_value)) +
  geom_point() +
  facet_grid(household_size ~ marital_status)
```  

### Knowledge check

```{block, type='todo'}
1. Filter for all transactions in the meat `department` and create a scatter plot comparing `quantity` to `sales_value`.
2. Build onto the plot in #1 by faceting by `product_category` using `facet_wrap(~product_category)`. What does this plot tell you?
3. Build onto the plot in #1 by faceting by `product_category` and `brand` using `facet_grid(product_category~brand)`. What does this plot tell you? What's the difference between `facet_wrap` and `facet_grid`?
```

## Labels & Annotations

Textual labels and annotations (on the plot, axes, geometry, and legend) are an important part of making a plot understandable and communicating information. Although not an explicit part of the *Grammar of Graphics* (the would be considered a form of geometry), `ggplot` makes it easy to add such annotations.

You can add titles and axis labels to a chart using the `labs()` function (not `labels`, which is a different R function!):

```{r basic-graph-labels-titles}
hshld_by_basket <- df %>%
   group_by(basket_id, household_size) %>%
   summarize(quantity = sum(quantity), sales_value = sum(sales_value))

ggplot(hshld_by_basket, aes(x = quantity, y = sales_value, color = household_size)) +
   geom_point() +
   labs(title = "Customer Net Spend by Basket",
        subtitle = "Household transaction data covering 2016-2017.",
        x = "Quantity of items in each basket",
        y = "Total net spend per basket ($)",
        color = "Household size")
```

It is also possible to add labels into the plot itself (e.g., to label each point or line) by adding a new `geom_text` or `geom_label` to the plot; effectively, you’re plotting an extra set of data which happen to be the variable names:

```{r label-in-ggplot, execute=TRUE, message=FALSE, fig.width=8}
top_baskets <- hshld_by_basket %>%
   ungroup() %>%
   slice_max(sales_value, n = 10)

ggplot(hshld_by_basket, aes(x = quantity, y = sales_value)) + 
  geom_point(aes(color = household_size)) +
  geom_label(data = top_baskets, aes(label = basket_id))
``` 
  
However, note that many labels overlap one-another in the center of the plot and some labels hang off the page. We can use the `geom_label_repel` function from the `ggrepel` package to help position labels.

```{r label-ggplot-no-overlap, execute=TRUE, message=FALSE, fig.width=8}
library(ggrepel)

ggplot(hshld_by_basket, aes(x = quantity, y = sales_value)) + 
  geom_point(aes(color = household_size)) +
  geom_label_repel(data = top_baskets, aes(label = basket_id))
```  

### Knowledge check

```{r}
# exercise data
hshld_df <- df %>%
   group_by(household_id) %>%
   summarize(
      total_sales = sum(sales_value),
      total_quantity = sum(quantity)
   ) %>%
   mutate(dollar_per_item = total_sales / total_quantity)
```

```{block, type='todo'}
Using the data above:

1. Create a bar plot of the top 10 households that spend the most per item purchased?  Refine this plot by adding a title and better x and y-axis names.
2. Create a scatter plot that compares `total_quantity` to `total_sales`. Log transform the x and y axis. Refine this plot by adding a title and better x and y-axis names.
```

## Exercises

```{block, type='todo'}
1. Identify all different products that contain “pizza” in their `product_type` description. 
2. Use a bar plot to assess whether married versus unmarried customers purchase more pizza products.
3. Create a scatter plot to assess the `quantity` versus `sales_value` of pizza products sold.
4. Now facet the scatter plot in #3 by `product_category`. Do the results surprise you?
```


## Additional Resources on **ggplot2**

This gets you started with **ggplot2**; however, this a lot more to learn. The following resources provide additional avenues to learn more:

* [gglot2 Documentation](https://ggplot2.tidyverse.org/) (particularly the [function](https://ggplot2.tidyverse.org/reference/index.html) reference)
* [ggplot2 Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf) (see also [here](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/))
* [Data Visualization portion of R for Data Science Book](http://r4ds.had.co.nz/data-visualisation.html)
* [ggplot2 book](https://ggplot2-book.org/)
* [A Layered Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)
* [ggplot2 extensions](https://exts.ggplot2.tidyverse.org/) that will give you even more capabilities such as interactive plots, mapping, etc!
